
#**Reproducible Research - Peer Assessment 1**
***

This is the R Markdown document for peer assessment#1 of Coursera course - "Reproducible Research".

##**Expectations**
***
Using the steps data available in Github, the following steps are required to be carried out.


1. Code for reading in the dataset and/or processing the data
2. Histogram of the total number of steps taken each day
3. Mean and median number of steps taken each day
4. Time series plot of the average number of steps taken
5. The 5-minute interval that, on average, contains the maximum number of steps
6. Code to describe and show a strategy for imputing missing data
7. Histogram of the total number of steps taken each day after missing values are imputed
8. Panel plot comparing the average number of steps taken per 5-minute interval across weekdays and weekends
9. All of the R code needed to reproduce the results (numbers, plots, etc.) in the report

###**1. Get the files and load the data**

Load the libraries

```{r, eval=TRUE, include=TRUE, echo=TRUE}
#Load libraries
library(ggplot2)
library(lattice)
```
Unzip the zip file and load the csv data
```{r, eval=TRUE, include=TRUE, echo=TRUE}
# Get the data files
if(!file.exists("activity.csv"))
{
	unzip(zipfile="repdata-data-activity.zip") 
}
#Read the data
stepsData <- read.csv("activity.csv")

```
###**2. Histogram of the total number of steps taken each day**
Using aggregate function, summarize by date and plot histogram 

```{r, eval=TRUE, include=TRUE, echo=TRUE}
# Summarize the steps data by day and plot histogram
stepsDataByDate <- aggregate(steps ~ date, data=stepsData, FUN=sum,na.rm=TRUE)
hist(stepsDataByDate$steps,col = "blue",main = "Histogram of the total number of daily steps",xlab = "Total number of steps taken each day",breaks = 20)
```

###**3. Mean and median number of steps**
Compute and display mean and median 

```{r, eval=TRUE, include=TRUE, echo=TRUE}
#Mean and median
stepsMean <- mean(stepsDataByDate$steps)
stepsMedian <- median(stepsDataByDate$steps)
stepsMean
stepsMedian
```

###**4. Time series plot of the average number of steps taken**
Using aggregate function, summarize by interval and plot 
```{r, eval=TRUE, include=TRUE, echo=TRUE}
#Time series Plot
stepsIntervalData <- aggregate(steps ~ interval, data=stepsData, FUN = mean, na.rm=TRUE)
plot2 <- ggplot(stepsIntervalData, aes(interval, steps)) + geom_line(color = "blue") + labs(title = "Time Series Plot of the 5-minute Intervals", x = "Intervals (5 minute)", y = "Average number of steps taken")
print(plot2)
```

###**5. The 5-minute interval that contains the maximum number of steps**
Using which.max, display the value
```{r, eval=TRUE, include=TRUE, echo=TRUE}
stepsIntervalData$interval[which.max(stepsIntervalData$steps)]
```

###**6. Code to describe and show a strategy for imputing missing data**
Here we use the interval mean to replace the missing values
```{r, eval=TRUE, include=TRUE, echo=TRUE}
#Calculate number of missing values
sum(is.na(stepsData))

#Fill missing values with interval mean
stepsDataImputed <- stepsData
for (i in 1:nrow(stepsDataImputed)) 
{
    if (is.na(stepsDataImputed$steps[i])) 
        stepsDataImputed$steps[i] <- stepsIntervalData[which(stepsDataImputed$interval[i] == stepsIntervalData$interval), ]$steps
}

#create new dataset
write.csv(stepsDataImputed, "activity1.csv")
```
###**7. Histogram of the total daily steps after imputing missing values**
Repeat the same as before (i.e., summarize by date and plot historgram) with the revised data.

```{r, eval=TRUE, include=TRUE, echo=TRUE}
# Summarize the steps data by day and plot histogram with new data (imputed)
stepsDataByDateNew <- aggregate(steps ~ date, data=stepsDataImputed, FUN=sum, na.rm=TRUE)
hist(stepsDataByDateNew$steps,col = "blue",main = "Histogram of the total number of daily steps (Imputed)",xlab = "Total number of steps taken each day",breaks = 20)
```
Compute revised mean and median
```{r, eval=TRUE, include=TRUE, echo=TRUE}
#Mean and median with new data (imputed)
stepsMeanNew <- mean(stepsDataByDateNew$steps)
stepsMedianNew <- median(stepsDataByDateNew$steps)
stepsMeanNew 
stepsMedianNew 
```
Based on both of these, the imputing of interval means for missing values has not changed the means and histogram much.

###**8. Comparing daily steps across weekdays and weekends**

First create a function to tell a daty if it is weekday or weekend
```{r, eval=TRUE, include=TRUE, echo=TRUE}
#Differences between Weekday and Weekends
#Function to identify daytype
daytype <- function(inputDate) 
{
    if (weekdays(as.Date(inputDate)) %in% c("Saturday", "Sunday")) 
       "Weekend"
    else 
       "Weekday"
    
}
```
Next apply it for all the rows, so we know if a given date is weekday or weekend
```{r, eval=TRUE, include=TRUE, echo=TRUE}
#Add Weekday/Weekend information to data
stepsDataImputed$daytype <- sapply(stepsDataImputed$date, daytype)
```
Now aggregate across interval and day type, and plot
```{r, eval=TRUE, include=TRUE, echo=TRUE}
# Summarize the steps data by day and plot with new data (imputed)
stepsDataByDayType <- aggregate(stepsDataImputed$steps, list(stepsDataImputed$interval, stepsDataImputed$daytype),FUN = "mean")
names(stepsDataByDayType) <- c("interval","daytype","mean.steps")
plot4 <- xyplot(stepsDataByDayType$mean.steps ~ stepsDataByDayType$interval | stepsDataByDayType$daytype, layout = c(1, 2), type = "l", xlab = "Interval", ylab = "Mean number of steps")
print(plot4)
```
